---
title: "Fluent genomic data analyses with plyranges and tximeta"
author: "Michael Love, Michael Lawrence, Stuart Lee"
---

# RNA-seq data import

```{r}
suppressPackageStartupMessages(library(plyranges))
library(macrophage)
dir <- system.file("extdata", package="macrophage")
coldata <- file.path(dir, "coldata.csv") %>% 
  read.csv() %>% 
  select(
    names, 
    id = sample_id, 
    line = line_id, 
    condition = condition_name
  ) %>%
  mutate(
    files = file.path(dir, "quants", names, "quant.sf.gz"),
    condition = relevel(condition, "naive")
  )
head(coldata)
```

```{r}
suppressPackageStartupMessages(library(SummarizedExperiment))
library(tximeta)
```

```{r}
se <- tximeta(coldata, dropInfReps=TRUE)
assayNames(se)
gse <- summarizeToGene(se)
```

# Basic DE task

```{r}
library(DESeq2)
dds <- DESeqDataSet(gse, ~line + condition)
keep <- rowSums(counts(dds) >= 10) >= 6
dds <- dds[keep,]
```

```{r}
dds <- DESeq(dds)
res <- results(dds, contrast=c("condition","IFNg","naive"),
               lfcThreshold=1)
summary(res, alpha=.01)
DESeq2::plotMA(res, ylim=c(-10,10))
plotCounts(dds, which.min(res$pvalue), "condition")
```

Output *GRanges* results:

```{r}
de_genes <- results(dds, 
                    contrast=c("condition","IFNg","naive"),
                    lfcThreshold=1, 
                    format="GRanges") %>% 
  filter(padj < 0.01) %>%
  mutate(de_log2FC = log2FoldChange,
         de_padj = padj)
```

```{r}
# how to do this with plyranges?
mcols(de_genes) <- mcols(de_genes)[c("de_log2FC","de_padj")]
```


# ATAC peak data

<https://zenodo.org/record/1188300#.XIAhXlNKjOQ>

This takes ~30 seconds and loads an object of ~370 Mb.

```{r}
atac_mat <- as.matrix(read.delim("ATAC_cqn_matrix.txt.gz"))
```

```{r}
atac_coldata <- read.delim("ATAC_sample_metadata.txt.gz") %>% 
  # AnnotationDbi overrides select
  plyranges::select(
    sample_id,
    donor,
    condition = condition_name
  ) %>% 
  mutate(condition = relevel(condition, "naive"))
```

```{r}
library(GenomicRanges)
peaks_df <- read.delim("ATAC_peak_metadata.txt.gz")
# add the genome info
peaks_gr <- peaks_df %>% 
  plyranges::select(seqnames = chr, start, end, gene_id) %>% 
  as_granges()
```

```{r}
idx <- match(colnames(atac_mat), atac_coldata$sample_id)
atac_coldata <- atac_coldata[idx,]
all.equal(colnames(atac_mat), as.character(atac_coldata$sample_id))
```

```{r}
atac <- SummarizedExperiment(list(cqndata=atac_mat),
                             rowRanges=peaks_gr,
                             colData=atac_coldata)
```

```{r}
rmu <- rowMeans(assay(atac))
rvar <- rowVars(assay(atac))
idx <- sample(nrow(atac),1e4)
plot(rmu[idx], rvar[idx], cex=.1)
```

```{r}
library(limma)
design <- model.matrix(~donor + condition, colData(atac))
fit <- lmFit(assay(atac), design)
fit <- eBayes(fit)
idx <- which(colnames(fit$coefficients) == "conditionIFNg")
tt <- topTable(fit, coef=idx, sort.by="none", n=nrow(atac))
```

```{r}
idx <- which.max(tt$logFC)
plot(assay(atac)[idx,] ~ atac$condition)
table(tt$logFC > 1 & tt$adj.P.Val < .05)
```

```{r}
peaks <- rowRanges(atac) %>% 
  mutate(
    da_logFC = tt$logFC,
    da_padj = tt$adj.P.Val
  )
seqlevelsStyle(peaks) <- "UCSC"
```

# plyranges

```{r}
suppressPackageStartupMessages(library(plyranges))
```

```{r}
da_peaks <- peaks %>% filter(da_logFC > 2 & da_padj < .01)
other_peaks <- peaks %>% filter(da_padj > .1)
```

```{r}
# sub-sample, here define a dplyr function first
sample_n <- function(x,n) x[sample.int(length(x),n,replace=FALSE)]
other_peaks <- other_peaks %>%
  sample_n(length(da_peaks))
```

```{r}
all_peaks <- bind_ranges(da=da_peaks,
                         not_da=other_peaks,
                         .id="origin") %>%
  mutate(
    # clean up the origin column
    origin = sub("da.*", "da", origin)
  )
``` 

```{r}
overlap_peaks <- all_peaks %>%
  mutate(width=1e4) %>%
  join_overlap_left(de_genes)
```

```{r}
overlap_peaks %>%
  group_by(origin) %>%
  summarize(close=sum(!is.na(de_padj)))
```

At this point, one can adjust the DE thresholds, but it would be more
complicated to adjust the peak thresholds, because of how the
subsampling was done to balance the two groups. So we could drop the
subsampling and instead look at the proportion of overlap, which would
allow one to adjust both DE and DA thresholds and compare overlap.

```{r}
overlap_peaks %>%
  group_by(origin) %>%
  summarize(close=sum(!is.na(de_padj) & abs(de_log2FC) > 2))
```

```{r}
overlap_peaks %>%
  group_by(origin) %>%
  summarize(close=sum(!is.na(de_padj) & abs(de_log2FC) > 3))
```
