---
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    toc_depth: 3
    fig_width: 5
bibliography: "`r file.path(system.file(package='plyrangesTximetaCaseStudy', 'vignettes'), 'library.bib')`"
vignette: >
  %\VignetteIndexEntry{plyrangesTximetaCaseStudy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding[utf8]{inputenc}
---

# A plyranges and tximeta case study: Overlapping differential expression and differential chromatin accessibility

## Instructor(s) name(s) and contact information

[Michael Love](https://mikelove.github.io),
[Stuart Lee](https://stuartlee.org/),
[Michael Lawrence](https://twitter.com/lawremi)

## Introduction

In this case study, we will examine a subset of the RNA-seq and
ATAC-seq data from @alasoo -
[doi: 10.1038/s41588-018-0046-7](https://doi.org/10.1038/s41588-018-0046-7). 
The experiment involved treatment of macrophage cell lines from a number
of human donors with IFN gamma, *Salmonella* infection, or both
treatments combined. In the original study, the authors examined gene
expression and chromatin accessibility in a subset of 86
successfully differentiated iPSC lines, and examined baseline QTL and
response QTL for both expression and accessibility. Here we will not
examine the genotypes, but only the expression data and chromatin
accessibility data, which is publicly available. First, let us review
the main finding as presented in the abstract:

> We observed that ~60% of stimulus-specific expression QTLs with a
> detectable effect on chromatin altered the chromatin accessibility
> in naive cells, thus suggesting that they perturb enhancer
> priming. Such variants probably influence binding of
> cell-type-specific transcription factors, such as PU.1, which can
> then indirectly alter the binding of stimulus-specific transcription
> factors, such as NF-κB or STAT2. Thus, although chromatin
> accessibility assays are powerful for fine-mapping causal regulatory
> variants, detecting their downstream effects on gene expression will
> be challenging, requiring profiling of large numbers of stimulated
> cellular states and time points.

**Scientific aim of the case study**: here we will perform a much simpler
analysis. We will examine the effect of IFNg stimulation on gene
expression and chromatin accessibility, and look to see if there is an
enrichment of differentially accessible (DA) ATAC-seq peaks in the
vicinity of differentially expressed (DE) genes. This is plausible, as
the transcriptomic response to IFNg stimulation may be mediated
through binding of regulatory proteins to accessible regions, and this
binding may increase the accessibility of those regions such that it
can be detected by ATAC-seq.

**Technical aim of the case study**: we will show how to use the
Bioconductor package *plyranges* and *tximeta*. The first package can
be used to perform easily-readable manipulations of data tied to genomic
ranges, e.g. shifts, windows, overlaps, etc. The *plyranges* package
is described by @Lee2019, and leverages underlying range operations
described by @granges. The second package is
used to read RNA-seq quantification into R/Bioconductor, such that the
genomic ranges are automatically attached to the quantification data
and differential expression results. 

We begin by loading the RNA-seq data contained in the Bioconductor
Experiment Data package, *macrophage*. The package contains RNA-seq
quantification from 24 RNA-seq samples, a subset of the
RNA-seq samples generated and analyzed by @alasoo. The paired-end
reads were quantified using *Salmon* [@salmon], using the Gencode 29
human reference transcripts [@gencode]. For more details on
quantification, and the exact code used, consult the vignette of the
[macrophage](http://bioconductor.org/packages/macrophage) package.
The package also contains the `Snakemake` file that was used to
distribute the *Salmon* quantification jobs on a cluster
[@snakemake].

## RNA-seq data import with *tximeta*

First, we specify a directory `dir`, where the quantification files
are stored. *You typically would not use* `system.file` *to specify
this directory.* You could simply specify this directory with:

```{r eval=FALSE}
dir <- "/path/to/quant/files"
```

However, here we have distributed the files in an R package, so we
locate that R package and the files using `system.file`.

The rest of the chunk reads in the `coldata.csv` file, selects
relevant columns from this table, and created `files` and
`condition` columns which point to the quantification files and
specify the condition (as a factor). One last detail is that the
quantification files have been gzipped, so they have the `.gz`
ending. Typically, these would look like `quant.sf` without the `.gz`
ending. 

```{r}
suppressPackageStartupMessages(library(plyranges))
library(macrophage)
dir <- system.file("extdata", package="macrophage")
coldata <- file.path(dir, "coldata.csv") %>% 
  read.csv() %>% 
  dplyr::select(
    names, 
    id = sample_id, 
    line = line_id, 
    condition = condition_name
  ) %>%
  mutate(
    files = file.path(dir, "quants", names, "quant.sf.gz"),
    condition = relevel(condition, "naive")
  )
head(coldata)
```

We next load the *SummarizedExperiment* and *tximeta* packages.

```{r}
suppressPackageStartupMessages(library(SummarizedExperiment))
library(tximeta)
```

The following lines of code do a lot of work for the user: importing
the RNA-seq quantification (dropping inferential replicates in this
case), locating the relevant reference transcriptome, attaching
the transcript ranges to the data, and fetching genome information. We
can see what data has been imported using `assayNames`, which lists
the assays of the *SummarizedExperiment*.

```{r}
se <- tximeta(coldata, dropInfReps=TRUE)
assayNames(se)
```

Because `tximeta` knows the correct reference transcriptome, we can
ask `tximeta` to summarize the transcript-level data to the gene level
using the methods of @Soneson2015.

```{r}
gse <- summarizeToGene(se)
```

## Basic RNA-seq DE analysis

We can easily run a differential expression analysis with *DESeq2*
using the following code chunks [@Love2014]. The design indicates that
we want to control for the donor (`line`) and test on the condition.

```{r}
library(DESeq2)
dds <- DESeqDataSet(gse, ~line + condition)
keep <- rowSums(counts(dds) >= 10) >= 6
dds <- dds[keep,]
```

```{r}
dds <- DESeq(dds)
res <- results(dds, contrast=c("condition","IFNg","naive"),
               lfcThreshold=1, alpha=0.01)
```

Always good to do a little visualization of results:

```{r}
summary(res)
DESeq2::plotMA(res, ylim=c(-10,10))
plotCounts(dds, which.min(res$pvalue), "condition")
```

We now output *GRanges* results (both DE and some non-DE for
comparison later). The `select` command pulls out particular columns
from the *GRanges*, in some cases renaming them so they will be unique
when we look at differential accessibility (which also will have LFC
and adjusted p-values).

```{r}
de_genes <- results(dds, 
                    contrast=c("condition","IFNg","naive"),
                    lfcThreshold=1, 
                    format="GRanges") %>%
  filter(padj < 0.01) %>%
  mutate(gene_id=names(.)) %>%
  plyranges::select(gene_id,
                    de_log2FC = log2FoldChange,
                    de_padj = padj)
```

We re-run `results` because we don't want to use an `lfcThreshold` 
this time, to pull out genes which are not differentially expressed
according to the *DESeq2* significance test.

```{r}
other_genes <- results(dds, 
                       contrast=c("condition","IFNg","naive"),
                       format="GRanges") %>% 
  filter(pvalue > 0.1) %>%
  mutate(gene_id=names(.)) %>%
  plyranges::select(gene_id,
                    de_log2FC = log2FoldChange,
                    de_padj = padj)
```

## ATAC-seq peak DA analysis

The ATAC-seq data can be downloaded from the following deposition URL:

<https://zenodo.org/record/1188300#.XIAhXlNKjOQ>

The data is fairly large. For running this section of workflow we need
the files:

* `ATAC_cqn_matrix.txt.gz` (109M)
* `ATAC_sample_metadata.txt.gz` (<1M)
* `ATAC_peak_metadata.txt.gz` (5.6M)

Due to the large size of the matrix file, we cannot bundle the data
with this workflow repository. We therefore set the following code
chunks to be unevaluated, until the creation of the `peaks` object.

The ATAC-seq data has already been normalized with *cqn* [@Hansen2012]
and log2 transformed. Loading the *cqn*-normalized matrix of log2
transformed read counts takes ~30 seconds and loads an object of ~370 Mb.

```{r eval=FALSE}
atac_mat <- as.matrix(read.delim("ATAC_cqn_matrix.txt.gz"))
```

We also read in the sample metadata:

```{r eval=FALSE}
atac_coldata <- read.delim("ATAC_sample_metadata.txt.gz") %>% 
  plyranges::select(
    sample_id,
    donor,
    condition = condition_name
  ) %>% 
  mutate(condition = relevel(condition, "naive"))
```

Finally, we read in the peak metadata (locations in the genome), and
convert to a *GRanges* object:

```{r eval=FALSE}
peaks_df <- read.delim("ATAC_peak_metadata.txt.gz", strings=FALSE)
peaks_gr <- peaks_df %>% 
  plyranges::select(seqnames = chr, start, end, peak_id=gene_id) %>% 
  as_granges()
# we know this from the Zenodo entry
# https://zenodo.org/record/1188300#.XJOFSlNKiL5
genome(peaks_gr) <- "GRCh38"
```

Now we make sure that our ATAC-seq quantifications and our metadata
are aligned:

```{r eval=FALSE}
idx <- match(colnames(atac_mat), atac_coldata$sample_id)
atac_coldata <- atac_coldata[idx,]
all.equal(colnames(atac_mat), as.character(atac_coldata$sample_id))
```

We then combine the data, and two pieces of metadata into a
*SummarizedExperiment*: 

```{r eval=FALSE}
atac <- SummarizedExperiment(list(cqndata=atac_mat),
                             rowRanges=peaks_gr,
                             colData=atac_coldata)
```

We can check the SD over mean plot, to assess for any systematic
trends:

```{r eval=FALSE}
rmu <- rowMeans(assay(atac))
rvar <- rowVars(assay(atac))
idx <- sample(nrow(atac),1e4)
plot(rmu[idx], sqrt(rvar[idx]), cex=.1)
```

For assessing differential accessibility, we run *limma* [@Smyth2004]: 

```{r eval=FALSE}
library(limma)
design <- model.matrix(~donor + condition, colData(atac))
fit <- lmFit(assay(atac), design)
fit <- eBayes(fit)
idx <- which(colnames(fit$coefficients) == "conditionIFNg")
tt <- topTable(fit, coef=idx, sort.by="none", n=nrow(atac))
```

We can plot the top gene by LFC:

```{r eval=FALSE}
idx <- which.max(tt$logFC)
plot(assay(atac)[idx,] ~ atac$condition)
table(tt$logFC > 1 & tt$adj.P.Val < .05)
```

We now take the `rowRanges` of the *SummarizedExperiment* and attach
the LFC and adjusted p-value from *limma*, so that we can consider the
overlap with differential expression.

```{r eval=FALSE}
peaks <- rowRanges(atac) %>% 
  mutate(
    da_log2FC = tt$logFC,
    da_padj = tt$adj.P.Val
  )
seqlevelsStyle(peaks) <- "UCSC"
genome(peaks) <- "hg38"
```

Due to restrictions on the size of this workflow repository, we do not
evaluate the above chunks, but instead now load the `peaks` object
that was saved after following the code above.

```{r}
library(plyrangesTximetaCaseStudy)
data(atac_peaks)
peaks
```

We filter to the set of peaks with a nominal FDR bound of 1%.

```{r}
da_peaks <- peaks %>% filter(da_padj < .01)
```

## Overlap analysis with *plyranges*

Now that we have DE genes, genes without strong signal of DE, and DA
peaks, we can perform our original aim to assess the enrichment of DA
peaks near DE genes. We've already used *plyranges* a number of times
above, to `filter`, `mutate` and `select` on *GRanges* objects, but
now we will get into more complicated use cases.

We define a function to sample `n` ranges from a *GRanges* object `x`.

```{r}
# sub-sample, here define a dplyr function first
sample_n <- function(x,n) x[sample.int(length(x),n,replace=FALSE)]
```

We downsample the number of "other" genes to the number of DE genes: 

```{r}
other_genes <- other_genes %>%
  sample_n(length(de_genes))
```

We then bind these two sets of ranges together and add a new column
`origin`. From `?bind_ranges`:

> When `.id` is supplied a new column is created that links each row to the original Range object. 

```{r}
lvls <- c("not_de","de")
all_genes <- bind_ranges(de=de_genes,
                         not_de=other_genes,
                         .id="origin") %>%
  mutate(
    origin = factor(origin, lvls)
  )
all_genes
``` 

We expand 10kb around the TSS of the genes using `mutate` with
`width=1e4`, and then perform a left join with the DA peaks. This
gives us the `all_genes` ranges (potentially with duplication), but
with the metadata columns from those overlapping DA peaks. For any
gene that has no overlaps, the DA peak columns will have `NA`'s.

```{r}
overlap_genes <- all_genes %>%
  mutate(width=1e4) %>%
  join_overlap_left(da_peaks)
overlap_genes
```

Now we can ask, how many DA peaks are near DE genes relative to
"other" genes? Note that a gene may appear more than once, because we
performed a left join. 

We group by DE vs "other", and sum up the DA peaks at any LFC, or
requiring an LFC of greater than 1, or greater than 2.

```{r}
overlap_tab <- overlap_genes %>%
  group_by(origin) %>%
  summarize(any=sum(!is.na(da_padj)),
            lfc1=sum(abs(da_log2FC) > 1, na.rm=TRUE),
            lfc2=sum(abs(da_log2FC) > 2, na.rm=TRUE))
overlap_tab
```

We can see that the enrichment increases for larger LFC threshold:

```{r}
sapply(overlap_tab[,-1], function(x) x[2]/x[1])
```

We can ask for the top peak per gene based on the peak's LFC, and pipe
that directly into a boxplot using *ggplot2*:

```{r}
library(ggplot2)
overlap_genes %>%
  group_by(gene_id) %>%
  plyranges::filter(abs(da_log2FC) == max(abs(da_log2FC))) %>%
  mcols %>% as.data.frame %>%
  ggplot(aes(origin, abs(da_log2FC))) +
  geom_boxplot()
```

## Further questions

* We specified a fixed-size window for computing overlaps. How could we
  have looked at various settings of this distance parameter?
* We computed the sum of DA peaks near the DE genes, for increasing
  LFC thresholds on the accessibility change. As we increased the
  threshold, the number of total peaks went down (likewise the mean
  number of DA peaks per gene). Likely the number of DE genes with a
  DA peak nearby with such a large change went down, but we didn't
  keep track of this. How could we have computed this? 
* We have some duplicate code in the `summarize` call, where we change
  only the LFC threshold. How could we perform this same operation
  over, say, a grid of 100 thresholds, without having to write the
  redundant code?
* How would you adapt the code to also consider variations on the
  thresholds applied to the DE genes (FDR 1%, testing against a null
  region of |LFC| < 1), and to the FDR cutoff (1%) for the DA peaks?
* We relied on the fact that the reference transcriptome was known to
  `tximeta`. What if this had not been the case?

## References
